Final Project Report: An exploration of Biliary Cirrhosis and Treatments

### 2023-12-14 CART.Rmd
b

```{=html}
<!-- -->
```
-   Class 1 (C): The patient is censored, meaning that the observation period ended without the death being recorded.

@@

```{r}
#Combine C and CL status into one variable and binarize
cirrhosisTreeData <- cirrhosis
cirrhosisTreeData$Status <- ifelse(cirrhosisTreeData$Status == "C" | cirrhosisTreeData$Status == "CL", 1, 0)
```

>>>>

-   Class 1 (C): The patient is censored, meaning that the observation period ended without the death being recorded.

@@

### Dataset for Tree
```{r}

cirrhosisTree <- cirrhosis

### 
```{r}

cirrhosisTree <- cirrhosisTree %>%
                                  filter(Status != "D") %>%
                                                      mutate(Transplant = ifelse(Status == 'C', 0, 1 ))

### 2023-12-14 EDA.Rmd
heyyy
b

---
title: "EDA"
author: "Mihir Kulkarni"
date: "2023-12-12"
output: html_document
---

@@

```{r}
library(reshape2)
library(ggplot2)
cirrhosis_melted <- melt(cirrhosis, id.vars = "Stage", measure.vars = c("Bilirubin", "Cholesterol", "Albumin", "Copper", "Alk_Phos", "SGOT"))
# Worry about , "Triglycerides" later
ggplot(cirrhosis_melted, aes(x = Stage, y = value, color = variable)) +
    geom_point() +
    scale_color_brewer(palette = "Dark2") +
    theme_minimal() +
    labs(title = "Biochemical Markers vs. Disease Stage", x = "Stage", y = "Marker Level")
```

### Gender Differences in Disease Progression: Is there a difference in the progression of PBC between male and female patients?

First, let us plot the sex distribution

```{r}
library(ggplot2)
library(dplyr)

@@

Investigating gender-based differences in disease progression can uncover any gender-specific patterns in PBC. This could lead to gender-tailored treatment approaches and a better understanding of the disease's biology, which might differ between males and females.

```{r}
surv_fit_gender <- survfit(surv_object ~ cirrhosis$Sex, data = cirrhosis)
ggsurvplot(surv_fit_gender, 
           data = cirrhosis,
           conf.int = TRUE, 
           palette = c("#FC4E07", "#2E9FDF"),
           xlab = "Days", 
           ylab = "Survival Probability", 
           title = "Survival Curves by Gender")

### Effect of Age at Diagnosis on Survival

@@

@@

Age at diagnosis is a critical factor in many diseases. This analysis can reveal if younger or older patients have better survival rates, influencing screening practices and early intervention strategies.

```{r}
ggplot(cirrhosis, aes(x = Age, y = N_Days, color = Status)) +
    geom_point() +
    scale_color_brewer(palette = "Set1") +
    theme_minimal() +
    labs(title = "Age at Diagnosis vs. Survival Time", x = "Age (days)", y = "Survival Time (days)")
```
### Comparison of Edema Status on Patient Outcomes

@@

```{r}
surv_fit_edema <- survfit(surv_object ~ cirrhosis$Edema, data = cirrhosis)
ggsurvplot(surv_fit_edema, 
           data = cirrhosis,
           conf.int = TRUE, 
           palette = c("#6ACC65", "#E7B800", "#00AFBB"),
           xlab = "Days", 
           ylab = "Survival Probability", 
           title = "Survival Curves by Edema Status")

>>>>

---
title: "EDA"
author: "Mihir Kulkarni, Nithika Menon"
date: "2023-12-12"
output: html_document
---

@@

### Sex Distribution of Patients
```{r}
library(ggplot2)
library(dplyr)

@@

```
Primary Biliary Cholangitis affects women much more than men at a 10-1 ratio. The data of this clinical study mimics the greater population. Investigating gender-based differences in disease progression can uncover any gender-specific patterns in PBC. This could lead to gender-tailored treatment approaches and a better understanding of the disease's biology, which might differ between males and females.

### Age Distribution of Patients

@@

```
The distribution of ages is quite normal.

### might have to mutate data below to filter out alive ppl but get the gist
```{r}

ggplot(data = cirrhosis, aes (Drug, Stage, fill = N_Days))  + geom_tile() + scale_fill_distiller(palette = "RdPu")

@@

### Individuals with Edema and if they were on the drug
```{r}

library(ggmosaic)
cirrhosis %>%
  ggplot() +
  geom_mosaic(aes( x = product(Edema), fill = Drug))
#ggplot(data = cirrhosis) + geom_mosaic(aes(x = Drug , fill = Status))

```
Looking at the dist. of stages
```{r}

cirrhosis %>%
      ggplot(aes(x = Stage)) + geom_bar() # need color

```{r}
cirrhosis %>%
      ggplot(aes(x = Cholesterol, y = Bilirubin)) + geom_line()

### Comparison of Edema Status on Patient Outcomes

@@

### 2023-12-14 Final Project.Rmd
Did some descriptiosn for the linear regression model
yuh
b

---
title: "Project #4: Final Project Report: An exploration of Cirrhosis"
output: html_notebook
---
hello

- Set the stage–explain the context you’re trying to learn something about. MUST ANSWER: Why is this important to you?

Cirrohosis is a condition where the liver is scarred and permanently damaged. As this condition, chances of liver failure and liver cancer increase. We are intrested in creating model that can predict stages of Cirrohosis and also death based on certain conditions. We hope our work can be useful in catching Cirrosis earlier on.

we're both interested in the health applications of AI and found an interesting article about cirrhosis and machine learning. We wanted to see if we could replicate the results and learn more about the disease.

- Explain the data provenance–Answer the Who, What, When, Where, Why, and How questions for the origin of the data. Include any additional information that would be useful in understanding the context in which the data are embedded.

@@

- Engage in Exploratory Data Analysis

@@

>>>>

---
title: "Final Project Report: An exploration of Biliary Cirrhosis and Treatments"
output: html_notebook
authors: "Mihir Kulkarni, Nithika Menon"
---

# Background

## What is Biliary Cholangitis?

Primary Biliary Cholangitis (previously called Primary Biliary Cirrhosis) is an autoimmune disease where bile ducts become swollen and inflamed and block the flow of bile. Bile is a substance that aids with digestion. The bile ducts carry bile from the liver to the small intestine. The swelling and inflammation can lead to scarring of the liver which is cirrhosis. Advanced cirrhosis can lead to liver failure or liver cancer. Medication can slow progression. There is no definite cure at this time.

## Analysis Objectives

### Q1: What predictors, if any, increases an individual's chance of receiving a liver transplant?

### Q2: Are there factors that can help distinguish different stages of Biliary Cholangitis not currently included in medical definition?

## Clinical Study + Dataset Background :

The dataset is produced from a clinical study by Mayo Clinic run from 1974-1984. The final 'Status' of each patient was observed in 1986. 'Status' was either Dead, Censored, or Censored due to liver transplant.

A link to the original study is here: https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf

@@

```{r}
count <- colSums(is.na(cirrhosis))

kable(count)
```
- Engage in Exploratory Data Analysis

@@

```{r}

cirrhosis %>% 
  ggplot(aes(x = N_Days)) + geom_histogram() +
  facet_wrap(facets = vars(Status))

```
